---
title: "Activity Monitoring Data"
output: html_document
---
## Loading and Processing 
```{r}
setwd("~/Documents/GitHub/Activity-Monitoring-Data")
if (!file.exists("activity.csv") )
    {
     dlurl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'  
     download.file(dlurl,destfile='repdata_data_activity.zip',mode='wb')  
     unzip('repdata_data_activity.zip')
    }

# Read data
data <- read.csv("activity.csv",sep=',', header=TRUE)
head(data)
```
### Calculate the total number of steps taken per day
The following code was used to collect the the number of daily steps between October and November, a 61-day period. A histogram is presented after the code demonstrating the frequency in step totals.
```{r}
daily<-c()  # Total number of daily steps 

for (i in 1:61){ # 61 days in period
    start<-(i-1)*288+1  # 288 five-minute steps in a day; 24*60/5=288
    last<-(i-1)*288+288
    temp<-data[start:last,1]    # extract 5-minute steps for each day
    daily<-c(daily,sum(temp))   # concatenate daily totals  
}

daily_noNA<-daily[!is.na(daily)]  # 8 NA's removed

hist(daily_noNA, xlab="steps",ylab="Frequency",col="grey",border="black",
     main="Total Number of Daily Steps Taken")
```
Here we reformat the dates into a more usable format.
```{r}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- factor(data$interval)
```
Here we create an index of non-missing entries
```{r}
NA_index <- is.na(as.character(data$steps))
data_no_NA <- data[!NA_index,]
head(data_no_NA)
```
The following histogram was created after removing empty data.
```{r}
steps_each_day <- aggregate(steps ~ date, data = data_no_NA, sum)
#Adding column names to the created data frame
colnames(steps_each_day) <- c("date", "steps")

hist(as.numeric(steps_each_day$steps), breaks = 20, col = "gray",border="black", xlab = "Number of Steps", main= "Total Number of Daily Steps")
```
Here we present the mean and median of the daily steps taken.
```{r}
mean(steps_each_day$steps)
median(steps_each_day$steps)
```
Here we calculate and plot the interval by the number of steps taken.
```{r}
#calculate the average
steps_per_interval <- aggregate(data_no_NA$steps, by=list(interval=data_no_NA$interval), FUN=mean)

#Add columns names
colnames(steps_per_interval) <- c("interval", "average_steps")

#plot the average daily activity pattern 
plot(as.integer(levels(steps_per_interval$interval)), steps_per_interval$average_steps, type="l",
     xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="black")
```
Here we code the maximum number of steps during an interval.
```{r}
#The maximum number of average steps
max_steps <- max(steps_per_interval$average_steps)
max_steps
```

```{r}
sum(is.na(as.character(data$steps)))
```

```{r}
sum(is.na(as.character(data$date)))
```

```{r}
sum(is.na(as.character(data$interval)))
```

```{r}
#finding the indices of missing values (NAs)
NA_index <- which(is.na(as.character(data$steps)))
complete_data <- data
#missing values using the mean for that 5-minute interval
complete_data[NA_index, ]$steps<-unlist(lapply(NA_index, FUN=function(NA_index){
                steps_per_interval[data[NA_index,]$interval==steps_per_interval$interval,]$average_steps
                }))
```

```{r}
#Checking the complete data with the summary and str methods
summary(complete_data)
```

```{r}
#Creating a data frame with the steps taken for each day
steps_each_day_complete <- aggregate(steps ~ date, data = complete_data, sum)
#Adding column names to the created data frame
colnames(steps_each_day_complete) <- c("date", "steps")

#Making the histogram
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "gray",border = "black", xlab = "Number of Steps", main= "Total Number of Daily Steps")
```
Here we calculate the mean and median steps in a full day.
```{r}
#Mean
mean(steps_each_day_complete$steps)
#Median
median(steps_each_day_complete$steps)
```
The following code will separate the time periods into Weekday and Weekend. We clearly see that activity levels are higher during the weekend.
```{r}
#Creating a factor variable "day "to store the day of the week:
complete_data$day <- as.factor(weekdays(complete_data$date))

#Creating a logical variable "is_weekday" (weekday=TRUE, weekend = FALE) :
complete_data$is_weekday <- ifelse(!(complete_data$day %in% c("Saturday","Sunday")), TRUE, FALSE) 


#Calculating the average number of steps for weekdays
weekdays_data <- complete_data[complete_data$is_weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)


#Calculating the average number of steps for weekends
weekends_data <- complete_data[!complete_data$is_weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)

#Adding columns names
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")
#Adding a column to indecate the day
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"

#Merging the two togather
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)
#Converting the day variabke to a factor
week_data$day <- as.factor(week_data$day)

#Making the plot
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, layout = c(1,2), type ="l", ylab="Number of Steps")
```